Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-06-04 paper/paper.tex
fix figure reference

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}. 
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
Observations such as these could help guide experimental studies to better understand protein function.

>>>>

To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).

However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}.
The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
Observations such as these could help guide experimental studies to better understand protein function.
